Making ice sheet models \\ scale properly

### 2023-01-07 talk/slides.tex
progress I hope

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item the geometry and velocity evolve \emph{in contact with its climate},
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt when floating (\emph{ice shelf})
    \item[$\circ$] calving into the ocean
    \end{itemize}
\item and the \emph{topography} on which it sits
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\medskip
\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})} for a model which does \emph{not} make these simplifications}
\item<2> {\footnotesize next, a basic definition of ``ice sheet model''}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
\end{definition} 
\item two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{balance of precipitation vs} \\
\text{melt \& runoff from surfaces}
\end{pmatrix}$$

@@

\bigskip
\item<2-> {\small map: \quad $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\begin{itemize}
\item nonlinear complementarity problem \onslide<2->{coupled to Stokes}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}}
\end{align*}

@@

\begin{itemize}
\item a thickness-transport form exposes the basic ice sheet model
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}
\item the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$
\item \emph{question:} is it really an advection?
\item[] \emph{answer:} not really, because ice flows (mostly) downhill
\item an alternative form is the \emph{shallow diffusion} equation for ice sheets:
   $$\frac{\partial H}{\partial t} = \nabla \cdot \left(D \grad s\right) + a$$
\item<2> {\footnotesize for more on the shallow diffusion perspective see Jouvet \& Bueler (2012)}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the model on the last slide is \emph{not yet} in common use for ice sheet simulations
    \begin{itemize}
    \item[$\circ$] for now, it is too expensive at scale
    \item[$\circ$] one goal: \alert{make it fast enough to use!}
    \end{itemize}
\item<2-> current-technology models use various shallow simplifications
\item<2-> simplest is a lubrication approximation, the \alert{shallow ice approximation}, in which $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator  
\end{itemize}
\end{frame}

\begin{frame}{shallow ice approximation: details}

@@

>>>>

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glacier geometry and velocity evolve \emph{in contact with its climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt when floating (\emph{ice shelf})
    \item[$\circ$] calving into the ocean
    \end{itemize}
\item and the \emph{topography} on which it sits
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\medskip
\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
\item<2> {\footnotesize next, a basic definition of ``ice sheet model''}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
\end{definition} 
\item two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

\smallskip
\item<2-> {\small map: \quad $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\begin{itemize}
\item nonlinear complementarity problem (NCP) \onslide<2->{coupled to Stokes}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}}
\end{align*}

@@

\begin{itemize}
\item a thickness-transport form exposes evolution type
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note: $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<2-> \emph{question:} is it really an advection?
\item<2->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\end{itemize}
\end{frame}

\begin{frame}{mass continuity equation: advection or diffusion?}

\begin{align*}
\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item both forms are highly-nonlinear: \, $\bU=\bU(H,\grad s), \, D = D(H,\grad s)$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the ``basic model'', i.e.~NCP coupled to Stokes, is \emph{not yet} in common use for ice sheet simulations
    \begin{itemize}
    \item[$\circ$] for now, it is too expensive at scale
    \item[$\circ$] goal: \alert{make it fast enough to use!}
    \end{itemize}

\smallskip
\item<2-> current-technology models use various shallow simplifications
\item<2-> simplest is a lubrication approximation, the \alert{shallow ice approximation}, in which $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator

\smallskip
\item<3-> the diffusion form of mass continuity ($\frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) = a$) is only literally valid in the lubrication approximation, but \emph{the fact that ice flows downhill has time-stepping stability consequences}
\end{itemize}
\end{frame}

\begin{frame}{shallow ice approximation NCP: details}

@@

\begin{frame}{time-stepping the mass continuity equation}

{\footnotesize
\begin{align*}
\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
}

\begin{itemize}
\item recall some traditional numerical analysis \dots
\end{itemize}
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item we expect \alert{explicit} time stepping for advections
\item for example, forward Euler in 1D gives
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bQ_{j+1/2}^\ell - \bQ_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] where $\bQ = \bU H$
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \end{itemize}
}
\only<2>{
\item we expect \alert{implicit} time stepping for diffusions
\item for example, backward Euler in 1D gives
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (H_{j+1}^{\ell+1} + H_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (H_{j}^{\ell+1} + H_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] or use: Crank-Nicolson, BDF, \dots
    \end{itemize}
}
\end{itemize}
\end{minipage}
\end{frame}

\begin{frame}{time-stepping in ice sheet models}

\begin{itemize}
\item current-technology models mostly ``believe'' that the problem is an advection and use explicit time stepping
\item FIXME but this has performance consequences
\item FIXME the problem is a DAE!
\item FIXME implicit alternative requires solving step equations: simultaneous NCP for SKE and Stokes
\end{itemize}
\end{frame}
